Exploring Local Explanations of Non-linear Models Using Animated Linear Projections

### 2022-01-08 _paper.tex
Saturday changes: - JMLR documents; .sty not effecting knit. - cheem scaling - Minor edits TODO: enlarge and clean up Case Studies & writing.

A comprehensive summary of the taxonomy and literature of explanation techniques is provided in Figure 6 of Arrieta et al. (2020). It includes a large number of model-specific explanations such as deepLIFT (Shrikumar et al. 2016; Shrikumar, Greenside, and Kundaje 2017), a popular recursive method for estimating importance in neural networks. There are fewer model-agnostic explanations, of which LIME (Ribeiro, Singh, and Guestrin 2016) SHAP (S. Lundberg and Lee 2017), and their variants are popular.

@@

SHaply Additive exPlanations (SHAP) approximates the variable importance in the vicinity of one observation conceptually by examining the effect of other variables on the contribution of the variable of interest on predicting the response. This explanations almost all point to Shapley (1953)'s method to evaluate an individual's contribution to cooperative games by permuting the players that contribute to the score. Strumbelj and Kononenko (2010) introduced the use of SHAP for local explanations in ML models. However, it is also related to partial dependence plots (Molnar 2020), used to explain the effect of a variable by predicting the response for a range of values on this variable, after fixing the value of all other variables to their mean. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in Wickham, Cook, and Hofmann (2015), which helps to understand the relative importance of each variable in the context of all other candidate variables.

For our application, we use \emph{tree SHAP}, a variant of SHAP enjoys a lower computational complexity (S. M. Lundberg, Erion, and Lee 2018). Tree SHAP is only compatible with tree-based models; we illustrate random forests. The following section will use normalized explanations as the starting projection basis (call this the \emph{attribution projection}) to further scrutinize the explanation.

@@

\caption{Illustration of the use of SHAP values for a random forest model for salaries of FIFA 2020 players based on nine predictors corresponding to different skills. Two observations (a star offensive player, Messi and a defensive player, van Dijk) are compared. Plot (a) shows the distribution of contributions for each variable across 25 permutations of predictors, with the median displayed as a dot, for both players. Plot b) is the break-down plot showing the SHAP values. It can be learned that reaction is important for explaining both players salaries. Offense and movement are important for Messi but not van Dijk, and conversely defense and power are important for van Dijk but not Messi.}\label{fig:shapdistrbd}
\end{figure}

@@

What information do we glean from using this method on a classification task? Typically we select a misclassified observation compared to a correctly classified point nearby in data space. The initial frame is the linear attribution of that observation's local explanation. By default, the manual tour varies the contribution of the variable with the largest difference between the primary and comparison observation; we can test the sensitivity of each variable to structure identified by the local explanation; we are exploring the support of the explanation, evaluating the support or robustness of the prediction.

@@

\caption{Overview of the cheem view for classification tasks. Plots are colored on predicted class, and red circles indicate misclassified observations. The radial tour is a 1D projection starting at the normalized tree SHAP values of the primary point. The first frame is the linear variable importances that best describe the observation's prediction. We probe the support of variable contributions by selecting a variable and use the radial tour to vary the coefficient.}\label{fig:classificationcase}
\end{figure}

@@

\begin{itemize}
    \item \textbf{Data:} a complete numerical matrix; explanatory and response variable. An optional categorical variable can be mapped to the color and shape of observations. Explanatory variables are scaled in visualization after modeling or creating local explanations. 
    \item \textbf{Model:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} [Liaw and Wiener (2002)] for compatibility with the local explanation, which requires tree-based models.
    \item \textbf{Local explanation:} any model-compatible linear explanation could be used. We apply tree SHAP, a more computationally efficient variant of SHAP, compatible with tree-based models. Tree SHAP was calculated with the package \textbf{treeshap} [Kominsarczyk et al. (2021), hosted on GitHub only]. The global view shows all observations in attribution space, requiring the variable importance from \emph{all} observations rather than just one.
\end{itemize}

The time to preprocess the data will vary significantly with the model and local explanation. For reference, the FIFA data, 5000 observations of nine explanatory variables, took 2.9 seconds to fit a random forest model of modest hyperparameters. Extracting the tree SHAP values of each observation took 254 seconds combined. PCA and statistics of the variables and attributions took 0.6 seconds. These runtimes were from a non-parallelized R session on a modern laptop, but suffice to say that the bulk of the time will be spent on the local attribution. An increase in model complexity or data dimensionality will quickly become an obstacle. With its reduced computational complexity, this makes tree SHAP a good candidate to start with. Alternatively, the package \textbf{fastshap} (Greenwell 2020) claims extremely low runtimes, which are attributed to fewer calls to the prediction function, partial implementation in C++, and efficient use of logical subsetting.

@@

\caption{Species classification of Palmer penguin data. We select a chinstrap penguin that is mislabeled as an adelie. By varying the contribution to bill length, we observe the explanation does not hold when bill length has a significant contribution. The .mp4 animation of this tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_penguins.mp4}}\label{fig:casepenguins}
\end{figure}

@@

\caption{Chocolates data type classification (milk or dark). We select a chocolate labeled as dark though a random forest model predicts it to be milk chocolate in light of the values on the nutritional label. We vary the contribution of calories from fat. Animated tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_chocolates.mp4}.}\label{fig:casechocolates}
\end{figure}

@@

\caption{FIFA 2020, regressing log wages {[}2020 Euros{]} from aggregations of skill measurements. The primary observation is a star offensive player (L. Messi) compared with a top defensive player (V. van Dijk). The animate radial tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_fifa.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_fifa.mp4}}\label{fig:casefifa}
\end{figure}

@@

\caption{Ames housing 2018 regressing log sales price {[}2018 USD{]}. Because the SHAP values are relatively well distributed across the variables, there is a lot of redundant information to explain the gap between select observations. This case may not be ideal for the cheem methodology, though we did find a near singular frame in the manual tour not, something we came looking for, but certainly, an exciting feature to find. The corresponding animation is at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_ames2018.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_ames2018.mp4}}\label{fig:caseames}
\end{figure}

@@

\leavevmode\vadjust pre{\hypertarget{ref-lundberg_unified_2017}{}}%
Lundberg, Scott, and Su-In Lee. 2017. {``A Unified Approach to Interpreting Model Predictions.''} \emph{arXiv Preprint arXiv:1705.07874}.

\leavevmode\vadjust pre{\hypertarget{ref-molnar_interpretable_2020}{}}%
Molnar, Christoph. 2020. \emph{Interpretable Machine Learning}. Lulu. com.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ribeiro_why_2016}{}}%
Ribeiro, Marco Tulio, Sameer Singh, and Carlos Guestrin. 2016. {``"{Why} {Should} {I} {Trust} {You}?": {Explaining} the {Predictions} of {Any} {Classifier}.''} \emph{arXiv:1602.04938 {[}Cs, Stat{]}}, February. \url{http://arxiv.org/abs/1602.04938}.

>>>>

A comprehensive summary of the taxonomy and literature of explanation techniques is provided in Figure 6 of Arrieta et al. (2020). It includes a large number of model-specific explanations such as deepLIFT (Shrikumar et al. 2016; Shrikumar, Greenside, and Kundaje 2017), a popular recursive method for estimating importance in neural networks. There are fewer model-agnostic explanations, of which LIME (Ribeiro, Singh, and Guestrin 2016) SHAP (Lundberg and Lee 2017), and their variants are popular.

@@

SHaply Additive exPlanations (SHAP) approximates the variable importance in the vicinity of one observation conceptually by examining the effect of other variables on the contribution of the variable of interest on predicting the response. This explanations almost all point to Shapley (1953)'s method to evaluate an individual's contribution to cooperative games by permuting the players that contribute to the score. Strumbelj and Kononenko (2010) introduced the use of SHAP for local explanations in ML models. While the models themselves are invariant to the order of the predictors, the additive explanations are not. The SHAP values are the mean contributions over some subset these permutations. This is also related to partial dependence plots (Molnar 2020), used to explain the effect of a variable by predicting the response for a range of values on this variable, after fixing the value of all other variables to their mean. Partial dependence plots are a global approximation of the variable importance, while SHAP is specific to one instance. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in Wickham, Cook, and Hofmann (2015), which helps to understand the relative importance of each variable in the context of all other candidate variables.

For our application, we use \emph{tree SHAP}, a variant of SHAP enjoys a lower computational complexity (Lundberg, Erion, and Lee 2018). Instead of aggregating over permutations of the variable order, tree SHAP approximates instance level feature importance by exploring the stucture of the decision trees. Tree SHAP is only compatible with tree-based models; we illustrate random forests. The following section will use normalized explanations as the starting projection basis (call this the \emph{attribution projection}) to further scrutinize the explanation.

@@

\caption{Illustration of the use of SHAP values for a random forest model for salaries of FIFA 2020 players based on nine predictors corresponding to different skills. A star offensive and densive player, Messi and a defensive player are compared, L. Messi and V van Dijk respectively. Plot a) is the break-down plot showing one additive variable-attributution. These attributions are sensitive to the order of the variables. Plot b) shows the distribution of attribution for each variable across 25 such permutation of the order of predictors, with the mean displayed as a dot, for each players. Offense and movement are important for Messi but not van Dijk, and conversely defense and power are important for van Dijk but not Messi.}\label{fig:shapdistrbd}
\end{figure}

@@

Typically we select a misclassified observation compared to a correctly classified point nearby in data space. The model infomation in the global view is a depiction of the model conmfusion matrix. The radial tour is 1D, with density display, while the goal of the tour; exploring the sensitivity of each variable to structure identified by the local explanation, evaluating the support or robustness of the prediction.

@@

\caption{The cheem view for classification tasks. Plots are colored on predicted class, and red circles indicate misclassified observations. Model visual is essentially a representation of the confusion matrix of the model. The radial tour is a 1D projection starting the attribution projection. The first frame is the linear variable importances that best describe the observation's prediction. We probe the support of variable contributions by selecting a variable and use the radial tour to vary its coefficient.}\label{fig:classificationcase}
\end{figure}

@@

\begin{itemize}
    \item \textbf{Data:} predictors are unscaled complete numerical matrix; while the response variable is occasionally log scaled for a more Gaussian distribution. Most models and local explanations will be scale invariant.
    \item \textbf{Model and explanation:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} [Liaw and Wiener (2002)], compatibility tree SHAP. We use modest hyperparameters, namely: 125 trees, number variables randomly sampled at each split, mtry = $\sqrt{p}$ or $p/3$ for classification and regression, and minimum size of terminal nodes $max(1, n/500)$ or $max(5, n/500)$ for classification and regression. Tree SHAP is calculated for \emph{each} observation using the package \textbf{treeshap} Kominsarczyk et al. (2021). This implementation aggregates over exhaustively over the attribution of all trees and we opt to not to fit interactions of features. 
    \item \textbf{Cheem view:} after the model and full explanation space are calculated we scale each feature by the standard deviation away from the mean to achieve a common support for visuals. Statistics are calculated on the scaled spaces. Interactive tabular display reports the original values.
\end{itemize}

The time to preprocess the data will vary significantly with the model and local explanation. For reference, the FIFA data, 5000 observations of nine explanatory variables, took 2.5 seconds to fit a random forest model of modest hyperparameters. Extracting the tree SHAP values of each observation took 270 seconds combined. PCA and statistics of the variables and attributions took 2.8 seconds. These runtimes were from a non-parallelized R session on a modern laptop, but suffice to say that the bulk of the time will be spent on the local attribution. An increase in model complexity or data dimensionality will quickly become an obstacle. With its reduced computational complexity, this makes tree SHAP a good candidate to start with. Alternatively, the package \textbf{fastshap} (Greenwell 2020) claims extremely low runtimes, which are attributed to fewer calls to the prediction function, partial implementation in C++, and efficient use of logical subsetting.

@@

\caption{Species classification of Palmer penguin data. We select a chinstrap penguin that is mislabeled as an adelie. By varying the contribution to bill length, we observe the explanation does not hold when bill length has a significant contribution. The .mp4 animation of this tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4}{shorturl.at/hFHNS}}\label{fig:casepenguins}
\end{figure}

@@

\caption{Chocolates data type classification (milk or dark). We select a chocolate labeled as dark though a random forest model predicts it to be milk chocolate in light of the values on the nutritional label. We vary the contribution of calories from fat. Animated tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4}{shorturl.at/dBE28}.}\label{fig:casechocolates}
\end{figure}

@@

\caption{FIFA 2020, regressing log wages {[}2020 Euros{]} from aggregations of skill measurements. The primary observation is a star offensive player (L. Messi) compared with a top defensive player (V. van Dijk). The animate radial tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_fifa.mp4}{shorturl.at/lCK26}}\label{fig:casefifa}
\end{figure}

@@

\caption{Ames housing 2018 regressing log sales price {[}2018 USD{]}. Because the SHAP values are relatively well distributed across the variables, there is a lot of redundant information to explain the gap between select observations. This case may not be ideal for the cheem methodology, though we did find a near singular frame in the manual tour not, something we came looking for, but certainly, an exciting feature to find. The corresponding animation is at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_ames2018.mp4}{shorturl.at/oqyDU}}\label{fig:caseames}
\end{figure}

@@

\leavevmode\vadjust pre{\hypertarget{ref-lundberg_unified_2017}{}}%
Lundberg, Scott M., and Su-In Lee. 2017. {``A Unified Approach to Interpreting Model Predictions.''} In \emph{Proceedings of the 31st International Conference on Neural Information Processing Systems}, 4768--77.

\leavevmode\vadjust pre{\hypertarget{ref-molnar_interpretable_2020}{}}%
Molnar, Christoph. 2020. \emph{Interpretable Machine Learning}. Lulu. com. \href{https://christophm.github.io/interpretable-ml-book/}{christophm.github.io/interpretable-ml-book/}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ribeiro_why_2016}{}}%
Ribeiro, Marco Tulio, Sameer Singh, and Carlos Guestrin. 2016. {``"{Why} {Should} {I} {Trust} {You}?": {Explaining} the {Predictions} of {Any} {Classifier}.''} In \emph{Proceedings of the 22nd {ACM} {SIGKDD} {International} {Conference} on {Knowledge} {Discovery} and {Data} {Mining}}, 1135--44. {KDD} '16. New York, NY, USA: Association for Computing Machinery. \url{https://doi.org/10.1145/2939672.2939778}.
